Rapid response of mesophyll conductance to light availability allows shade leaves to take advantage of sunflecks

### 2015-09-22 manuscript/manuscript_wtc3.Rmd
manuscript edits RD

# Summary
* Light gradients within tree canopies play a major role in the distribution of resources that define photosynthetic capacity of individual leaves. The transient nature of light availability within a canopy, however, complicates our ability to assess the contribution of shade leaves to canopy carbon gain.
* Leaf gas exchange was coupled with online carbon isotope discrimination to measure photosynthesis (A), stomatal conductance (g~s~) and mesophyll conductance (g~m~) of sun and shade leaves in *Eucalyptus tereticornis* trees grown in climate controlled whole tree chambers.
* More open stomata and rapid increases in mesophyll conductance

@@

## Whole tree chamber experimental design
Twelve *Eucalyptus tereticornis* seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, NSW, Australia. Each chamber has a height of 9 m and seedlings were grown for 15 months.  A detailed description of the WTC operation and design is available in [@barton2010whole]. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced an elevated air temperature treatment of +3&deg;C (ET, Figure `r airvarsfig`). Trees were watered weekly with 70 L from March 2013 to November 2013.  From December 2013 to final harvest trees were watered fortnightly with the mean monthly (100 yr) rainfall amount.  In February 2013 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Due to a limited range of data for the drought treatment only well-watered trees are reported, which reduces the sample size of WTC (n=6) after the initiation of this treatment.

@@

# Discussion
Here we show that leaves within canopies of *Eucalyptus tereticornis* trees maintain the ability to upregulate A beyond the limitations of their local light environment. More open stomata and rapid increases in g~m~ allowed shade leaves to readily utilize increases in light availability, leading to rates of leaf C gain exceeding sun leaves. Although these results should not be entirely unexpected as shade leaves experience interchanging periods of sun and shade [@pearcy1990sunflecks], a lack of empirical data within tree canopies still impedes our ability to predict whole canopy C gain. Failure to integrate this dynamic physiology into process based tree growth models may underestimate the ability of canopies to alter leaf C gain when environmental conditions change. Additionally, with measurements recorded across a large natural range of temperatures we were unable to detect any effects of a +3 &deg;C warming treatment on leaf physiology.

@@

\
Unexpected higher rates of g~s~ in shade leaves compared to sun leaves led to decreased ITE throughout the experiment. Additionally, consistently higher leaf $\delta$^13^C in shade leaves suggests that this pattern was prevalent across the leaf lifespan. From a canopy perspective this pattern in WUE initially appears to be detrimental to C gain as A in sun leaves was characterized by low rates of g~s~ and low C~i~.  As whole canopy C gain integrates the efficiency of all leaves, this begs the question of why shade leaves maintained a lower WUE when the photosynthetic efficiency of sun leaves appears at least partially constrained by water supply.

>>>>

# Summary
* Light gradients within tree canopies play a major role in the distribution of resources that define photosynthetic capacity of individual leaves. The transient nature of light availability within a canopy, however, complicates our ability to assess the contribution of shade leaves to canopy carbon gain.
* Leaf gas exchange was coupled with online carbon isotope discrimination to measure photosynthesis (A), stomatal conductance (g~s~) and mesophyll conductance (g~m~) of sun and shade leaves in *Eucalyptus tereticornis* trees grown in climate controlled whole tree chambers.
* Reduction in A, Vc~max, J~max and leaf nitrogen were detected in shade leaves, however, they maintained consistently higher g~s~ than sun leaves. When light intensity was increased for shade leaves g~s~ was higher in shade than in sun leaves, g~m~ increased compared to the low light environment and A in shade leaves was greater than sun leaves at the same PPFD.

@@

## Whole tree chamber experimental design
Twelve *Eucalyptus tereticornis* seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, NSW, Australia. Each chamber has a height of 9 m and seedlings were grown for 15 months.  A detailed description of the WTC operation and design is available in [@barton2010whole]. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced an elevated air temperature October 2013 to the end of the experiment trees were watered every 15 days with the mean monthly rainfall amount for Richmond, NSW.  In February 2013 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Due to a limited range of data for the drought treatment only well-watered trees are reported, which reduces the sample size of WTC (n=6) after the initiation of this treatment.

@@

# Discussion
Here we show that leaves within canopies of *Eucalyptus tereticornis* trees maintain the ability to upregulate A beyond the limitations of their local light environment. More open stomata and rapid increases in g~m~ allowed shade leaves to readily utilize increases in light availability, leading to rates of leaf C gain exceeding sun leaves. This finding is unexpected because although we know shade leaves experience interchanging periods of sun and shade [@pearcy1990sunflecks], a lack of empirical data within tree canopies still impedes our ability to predict whole canopy C gain. This dynamic behaviour of shade leaves complicates optimizations of canopy C uptake and accounting for the response of leaf physiology to transient light dynamics is an essential need in future studies. Additionally, with measurements recorded across a large natural range of temperatures we were unable to detect any effects of a +3 &deg;C warming treatment on leaf physiology.

@@

\
Unexpected higher rates of g~s~ in shade leaves compared to sun leaves led to decreased ITE throughout the experiment. Additionally, consistently higher leaf $\delta$^13^C in shade leaves suggests that this pattern was prevalent across the leaf lifespan. From a canopy perspective this pattern in water use efficiency initially appears to be detrimental to C gain as A in sun leaves was characterized by low rates of g~s~ and low C~i~. Relative to the differences in photosynthesis rate between leaf types the higher rates of g~s~ in shade leaves appear to exhibit waste water loss. As whole canopy C gain integrates the efficiency of all leaves, this begs the question of why shade leaves maintained a lower ITE.
